An Improved Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2021-07-20 simplified_sdc_intro/paper.tex
update to reflect the new formulation

\newcommand{\nse}[1]{{\mathrm{NSE}( #1 )}}

@@

We decompose the source terms, $\Sc(\Uc)$, into
hydrodynamical sources, $\Hb$ (like gravity), and reactive sources,
$\Rb$,
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc).
\end{equation}
The hydrodynamic sources include the ``pdV'' work for internal energy
as well as gravity
\begin{equation}
  \Hb(\Ub) = \left ( \begin{array}{c}
    0 \\
    0 \\
    0 \\
    \rho g \\
    \rho u g \\
    -p \partial u/\partial x \end{array} \right )
\end{equation}
 The reactive sources take the form:
\begin{equation}
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \omegadot(X_k) \\
     \rho \omegadot(\alpha_l) \\
     0 \\
     \rho \dot{S} \\
     \rho \dot{S} 
  \end{array} \right )
\end{equation}
where $\omegadot(X_k)$ is the creation rate for species $k$, $\omegadot(\alpha_l)$ is the creation rate for
auxiliary composition variable $l$, and $\dot{S}$
is the energy generation rate per unit mass.

@@

    We update the state by doing the integral in equation
    (\ref{eq:integral:simplesdc}).  Since we are approximating the
    advective term as piecewise constant in time, we can simply use an
    ODE integrator to integrate this, just as we do with the reaction
    system in Strang splitting.  The difference here being that we are
    integrating the conserved variables and the state sees the effect
    of advection as we integrate the reactions.  So rather than use
    $d\Uc/dt = \Rb(\Uc)$, the ODE form we use is
    \begin{equation}
      \odt{\Uc} = \Adv{\Uc}^{n+1/2,(k)} + \Rb(\Uc)
    \end{equation}
    This integration begins with $\Uc^n$ and results in $\Uc^{n+1,(k)}$.

    As we are integrating this system we need to get the
    temperature, $T$, for the rate evaluations.  We construct this
    by subtracting the kinetic energy from the total energy to get
    the specific internal energy, $e$, and then calling the equation
    of state.

    Our integrator also needs the Jacobian of the system, in terms of
    the conserved variables.  This is different than the form of the
    Jacobian usually used in reaction networks.  We describe the form
    of the Jacobian in appendix \ref{sec:app:jacobian}.

    Note that our advection terms are piecewise
    constant \MarginPar{need to think about this more} approximations,
    so when we integrate our system, their contributions will be
    linear in time.  Since $(\rho X_k)$ sees a linear advective term,
    $\rho$ does as well.  $(\rho u)$ will also have linear
    contributions in time from advection.  Since the reactions don't
    affect $\rho$ and $(\rho u)$, we can algebraically update these
    using this piecewise linear behavior:
    \begin{align}
      \rho(t) &= \rho^n + \Advss{\rho}^{n+1/2,(k)} \, t \\
      (\rho u)(t) &= (\rho u)^n + \Advss{\rho u}^{n+1/2,(k)} \, t
    \end{align}

@@

    \begin{itemize}
    \item Integrate the full reaction network (Eqs.~\ref{eq:strang:T} and \ref{eq:strang:X}) as usual for Strang splitting
    \item Update the aux quantities at the end of the burn using Eqs.~\ref{eq:aux:ye}, \ref{eq:aux:abar}, and \ref{eq:aux:bea} with the new mass fractions, $X_k$.
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:react_convert_strang} Convergence ($L_1$ norm) for the reacting convergence problem using SDC integration (2 iterations).}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $2.795 \times 10^{18}$  & 2.044  & $6.777 \times 10^{17}$  & 2.554  & $1.154 \times 10^{17}$  \\
 $\rho u$                    & $6.798 \times 10^{26}$  & 2.449  & $1.245 \times 10^{26}$  & 2.889  & $1.680 \times 10^{25}$  \\
 $\rho v$                    & $6.798 \times 10^{26}$  & 2.449  & $1.245 \times 10^{26}$  & 2.889  & $1.680 \times 10^{25}$  \\
 $\rho E$                    & $2.452 \times 10^{35}$  & 2.352  & $4.804 \times 10^{34}$  & 2.742  & $7.179 \times 10^{33}$  \\
 $\rho e$                    & $2.262 \times 10^{35}$  & 2.321  & $4.527 \times 10^{34}$  & 2.822  & $6.404 \times 10^{33}$  \\
 $T$                         & $2.236 \times 10^{21}$  & 1.691  & $6.927 \times 10^{20}$  & 2.481  & $1.240 \times 10^{20}$  \\
 $\rho X(\isotm{He}{4})$     & $2.879 \times 10^{18}$  & 2.020  & $7.096 \times 10^{17}$  & 2.529  & $1.229 \times 10^{17}$  \\
 $\rho X(\isotm{C}{12})$     & $1.697 \times 10^{17}$  & 1.950  & $4.393 \times 10^{16}$  & 2.231  & $9.357 \times 10^{15}$  \\
 $\rho X(\isotm{O}{16})$     & $1.686 \times 10^{14}$  & 1.659  & $5.338 \times 10^{13}$  & 1.957  & $1.375 \times 10^{13}$  \\
 $\rho X(\isotm{Fe}{56})$    & $2.795 \times 10^{8}$   & 2.044  & $6.777 \times 10^{7}$   & 2.554  & $1.154 \times 10^{7}$   \\
\enddata
\end{deluxetable}

@@

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:nse_strang_methodII} Convergence ($L_1$ norm) for the NSE convergence
problem using the simplified-SDC algorithm.}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $4.553 \times 10^{19}$  & 1.636  & $1.465 \times 10^{19}$  & 1.583  & $4.891 \times 10^{18}$  \\
 $\rho u$                    & $1.949 \times 10^{28}$  & 1.650  & $6.210 \times 10^{27}$  & 1.513  & $2.176 \times 10^{27}$  \\
 $\rho v$                    & $1.949 \times 10^{28}$  & 1.650  & $6.210 \times 10^{27}$  & 1.513  & $2.176 \times 10^{27}$  \\
 $\rho E$                    & $5.691 \times 10^{37}$  & 1.622  & $1.848 \times 10^{37}$  & 1.561  & $6.266 \times 10^{36}$  \\
 $\rho e$                    & $5.689 \times 10^{37}$  & 1.622  & $1.848 \times 10^{37}$  & 1.561  & $6.265 \times 10^{36}$  \\
 $T$                         & $5.778 \times 10^{20}$  & 1.740  & $1.729 \times 10^{20}$  & 1.702  & $5.316 \times 10^{19}$  \\
 $\rho X(\isotm{H}{1})$      & $4.553 \times 10^{-11}$ & 1.636  & $1.465 \times 10^{-11}$ & 1.583  & $4.891 \times 10^{-12}$ \\
 $\rho X(\isotm{He}{4})$     & $2.541 \times 10^{17}$  & 1.807  & $7.262 \times 10^{16}$  & 1.634  & $2.339 \times 10^{16}$  \\
 $\rho X(\isotm{C}{12})$     & $1.140 \times 10^{12}$  & 1.838  & $3.189 \times 10^{11}$  & 1.670  & $1.002 \times 10^{11}$  \\
 $\rho X(\isotm{O}{16})$     & $5.413 \times 10^{12}$  & 1.819  & $1.535 \times 10^{12}$  & 1.658  & $4.864 \times 10^{11}$  \\
 $\rho X(\isotm{Cr}{48})$    & $7.020 \times 10^{18}$  & 1.793  & $2.026 \times 10^{18}$  & 1.614  & $6.621 \times 10^{17}$  \\
 $\rho X(\isotm{Fe}{52})$    & $3.631 \times 10^{19}$  & 1.579  & $1.215 \times 10^{19}$  & 1.597  & $4.018 \times 10^{18}$  \\
 $\rho X(\isotm{Fe}{54})$    & $4.670 \times 10^{20}$  & 1.807  & $1.335 \times 10^{20}$  & 1.869  & $3.654 \times 10^{19}$  \\
 $\rho X(\isotm{Ni}{56})$    & $4.858 \times 10^{20}$  & 1.808  & $1.387 \times 10^{20}$  & 1.860  & $3.820 \times 10^{19}$  \\
 $\rho Y_e$                  & $2.372 \times 10^{19}$  & 1.649  & $7.564 \times 10^{18}$  & 1.587  & $2.518 \times 10^{18}$  \\
 $\rho \bar{A}$              & $4.009 \times 10^{21}$  & 1.693  & $1.240 \times 10^{21}$  & 1.690  & $3.843 \times 10^{20}$  \\
 $\rho (B/A)$                & $4.222 \times 10^{20}$  & 1.666  & $1.330 \times 10^{20}$  & 1.605  & $4.374 \times 10^{19}$  \\
\enddata
\end{deluxetable}

>>>>

\newcommand{\nse}[1]{{\mathtt{NSE}( #1 )}}

@@

We decompose the source terms, $\Sc(\Uc)$, into
hydrodynamical sources, $\Hb$ (like gravity), and reactive sources,
$\Rb$,
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc).
\end{equation}
The hydrodynamic sources include the ``pdV'' work for internal energy
as well as gravity
\begin{equation}
  \Hb(\Ub) = \left ( \begin{array}{c}
    0 \\
    0 \\
    0 \\
    \rho g \\
    \rho u g \\
    -p \partial u/\partial x \end{array} \right )
\end{equation}
 The reactive sources take the form:
\begin{equation}
  \label{eq:react_source}
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \omegadot(X_k) \\
     \rho \omegadot(\alpha_l) \\
     0 \\
     \rho \dot{S} \\
     \rho \dot{S} 
  \end{array} \right )
\end{equation}
where $\omegadot(X_k)$ is the creation rate for species $k$, $\omegadot(\alpha_l)$ is the creation rate for
auxiliary composition variable $l$, and $\dot{S}$
is the energy generation rate per unit mass.

@@

    We update the state by doing the integral in equation
    (\ref{eq:integral:simplesdc}).  Since we are approximating the
    advective term as piecewise constant in time, we can simply use an
    ODE integrator to integrate this, just as we do with the reaction
    system in Strang splitting.  The difference here being that we are
    integrating the conserved variables and the state sees the effect
    of advection as we integrate the reactions.  So rather than use
    $d\Uc/dt = \Rb(\Uc)$, the ODE form we use is
    \begin{equation}
      \odt{\Uc} = \Adv{\Uc}^{n+1/2,(k)} + \Rb(\Uc)
    \end{equation}

    Looking at the form of Eq.~\ref{eq:react_source}, we see that only
    the species, auxiliary quantities, and energies have source terms.
    Since we are reacting, we will use the $X_k$ as the primary
    composition variable and compute $\alpha_l$ as needed from them,
    so we don't directly integrate the auxiliary quantities.
    Likewise, the total and internal energies both provide the same
    information, and integrating both overconstrains the system, so we
    just integrate the internal energy.  We define the subset of variables
    that are directly integrated as:
    \begin{equation}
      \Uc^\prime = \left ( \begin{array}{c} \rho X_k \\ \rho e \end{array} \right )
    \end{equation}
    and we integrate
    \begin{equation}
      \odt{\Uc^\prime} = \Adv{\Uc^\prime}^{n+1/2,(k)} + \Rb(\Uc^\prime)
    \end{equation}
    This integration begins with $\Uc^{\prime,n}$ and results in $\Uc^{\prime,n+1,(k)}$.

    We will need the density at times during the integration, which we construct as:
    \begin{equation}
      \rho(t) = \rho^n + \Advss{\rho}^{n+1/2,(k)} \, t
    \end{equation}
    As we are integrating this system we need to get the temperature,
    $T$, for the rate evaluations.  We obtain this by directly from
    internal energy, composition, and density using the equation of
    state.

    Our integrator also needs the Jacobian of the system, in terms of
    the $\Uc^\prime$.  This is different than the form of the
    Jacobian usually used in reaction networks.  We describe the form
    of the Jacobian in appendix \ref{sec:app:jacobian}.

    Note that our advection terms are piecewise
    constant \MarginPar{need to think about this more} approximations,
    so when we integrate our system, their contributions will be
    linear in time.  Since $(\rho X_k)$ sees a linear advective term,
    $\rho$ does as well.  $(\rho u)$ will also have linear
    contributions in time from advection.

    At the end of the integration, we can do the conservative update of momentum
    and energy.  Momentum is straightforward, since there are no reactive sources:
    \begin{equation}
      (\rho u)^{n+1} = (\rho u)^n + \Delta t \Advss{\rho u}^{n+1/2,(k)}
    \end{equation}
    For total energy, we first need to isolate the reactive source for energy:
    \begin{equation}
      (\rho \Sdot)^{n+1/2} = \frac{(\rho e)^{n+1} - (\rho e)^n}{\Delta t} - \Advss{\rho e}^{n+1/2,(k)}
    \end{equation}
    Then the update is
    \begin{equation}
      (\rho E)^{n+1} = (\rho E)^n + \Delta t \Advss{\rho E}^{n+1/2,(k)} + \Delta t (\rho \Sdot)^{n+1/2}
    \end{equation}

@@

    \begin{itemize}
    \item Integrate the full reaction network (Eqs.~\ref{eq:strang:e} and \ref{eq:strang:X}) as usual for Strang splitting
    \item Update the aux quantities at the end of the burn using Eqs.~\ref{eq:aux:ye}, \ref{eq:aux:abar}, and \ref{eq:aux:bea} with the new mass fractions, $X_k$.
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:react_convert_sdc} Convergence ($L_1$ norm) for the reacting convergence problem using SDC integration (2 iterations).}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $2.795 \times 10^{18}$  & 2.044  & $6.777 \times 10^{17}$  & 2.554  & $1.154 \times 10^{17}$  \\
 $\rho u$                    & $6.798 \times 10^{26}$  & 2.449  & $1.245 \times 10^{26}$  & 2.889  & $1.680 \times 10^{25}$  \\
 $\rho v$                    & $6.798 \times 10^{26}$  & 2.449  & $1.245 \times 10^{26}$  & 2.889  & $1.680 \times 10^{25}$  \\
 $\rho E$                    & $2.452 \times 10^{35}$  & 2.352  & $4.804 \times 10^{34}$  & 2.742  & $7.179 \times 10^{33}$  \\
 $\rho e$                    & $2.262 \times 10^{35}$  & 2.321  & $4.527 \times 10^{34}$  & 2.822  & $6.404 \times 10^{33}$  \\
 $T$                         & $2.236 \times 10^{21}$  & 1.691  & $6.927 \times 10^{20}$  & 2.481  & $1.240 \times 10^{20}$  \\
 $\rho X(\isotm{He}{4})$     & $2.879 \times 10^{18}$  & 2.020  & $7.096 \times 10^{17}$  & 2.529  & $1.229 \times 10^{17}$  \\
 $\rho X(\isotm{C}{12})$     & $1.697 \times 10^{17}$  & 1.950  & $4.393 \times 10^{16}$  & 2.231  & $9.357 \times 10^{15}$  \\
 $\rho X(\isotm{O}{16})$     & $1.686 \times 10^{14}$  & 1.659  & $5.338 \times 10^{13}$  & 1.957  & $1.375 \times 10^{13}$  \\
 $\rho X(\isotm{Fe}{56})$    & $2.795 \times 10^{8}$   & 2.044  & $6.777 \times 10^{7}$   & 2.554  & $1.154 \times 10^{7}$   \\
\enddata
\end{deluxetable}

@@

\begin{deluxetable}{lllllll}
\tablecaption{\label{table:nse_sdc_methodII} Convergence ($L_1$ norm) for the NSE convergence
problem using the simplified-SDC algorithm.}
\tablehead{\colhead{field} & \colhead{$\epsilon_{64 \rightarrow 128}$} & 
           \colhead{rate} & \colhead{$\epsilon_{128\rightarrow 256}$} & 
           \colhead{rate} & \colhead{$\epsilon_{256\rightarrow 512}$}}
\startdata
 $\rho$                      & $4.553 \times 10^{19}$  & 1.636  & $1.465 \times 10^{19}$  & 1.583  & $4.891 \times 10^{18}$  \\
 $\rho u$                    & $1.949 \times 10^{28}$  & 1.650  & $6.210 \times 10^{27}$  & 1.513  & $2.176 \times 10^{27}$  \\
 $\rho v$                    & $1.949 \times 10^{28}$  & 1.650  & $6.210 \times 10^{27}$  & 1.513  & $2.176 \times 10^{27}$  \\
 $\rho E$                    & $5.691 \times 10^{37}$  & 1.622  & $1.848 \times 10^{37}$  & 1.561  & $6.266 \times 10^{36}$  \\
 $\rho e$                    & $5.689 \times 10^{37}$  & 1.622  & $1.848 \times 10^{37}$  & 1.561  & $6.265 \times 10^{36}$  \\
 $T$                         & $5.778 \times 10^{20}$  & 1.740  & $1.729 \times 10^{20}$  & 1.702  & $5.316 \times 10^{19}$  \\
 $\rho X(\isotm{H}{1})$      & $4.553 \times 10^{-11}$ & 1.636  & $1.465 \times 10^{-11}$ & 1.583  & $4.891 \times 10^{-12}$ \\
 $\rho X(\isotm{He}{4})$     & $2.541 \times 10^{17}$  & 1.807  & $7.262 \times 10^{16}$  & 1.634  & $2.339 \times 10^{16}$  \\
 $\rho X(\isotm{C}{12})$     & $1.140 \times 10^{12}$  & 1.838  & $3.189 \times 10^{11}$  & 1.670  & $1.002 \times 10^{11}$  \\
 $\rho X(\isotm{O}{16})$     & $5.413 \times 10^{12}$  & 1.819  & $1.535 \times 10^{12}$  & 1.658  & $4.864 \times 10^{11}$  \\
 $\rho X(\isotm{Cr}{48})$    & $7.020 \times 10^{18}$  & 1.793  & $2.026 \times 10^{18}$  & 1.614  & $6.621 \times 10^{17}$  \\
 $\rho X(\isotm{Fe}{52})$    & $3.631 \times 10^{19}$  & 1.579  & $1.215 \times 10^{19}$  & 1.597  & $4.018 \times 10^{18}$  \\
 $\rho X(\isotm{Fe}{54})$    & $4.670 \times 10^{20}$  & 1.807  & $1.335 \times 10^{20}$  & 1.869  & $3.654 \times 10^{19}$  \\
 $\rho X(\isotm{Ni}{56})$    & $4.858 \times 10^{20}$  & 1.808  & $1.387 \times 10^{20}$  & 1.860  & $3.820 \times 10^{19}$  \\
 $\rho Y_e$                  & $2.372 \times 10^{19}$  & 1.649  & $7.564 \times 10^{18}$  & 1.587  & $2.518 \times 10^{18}$  \\
 $\rho \bar{A}$              & $4.009 \times 10^{21}$  & 1.693  & $1.240 \times 10^{21}$  & 1.690  & $3.843 \times 10^{20}$  \\
 $\rho (B/A)$                & $4.222 \times 10^{20}$  & 1.666  & $1.330 \times 10^{20}$  & 1.605  & $4.374 \times 10^{19}$  \\
\enddata
\end{deluxetable}
